Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions

### 2021-04-02 Latex/appendix/webappendix.Rnw
Update text

\if1\blind{
\date{%
$^*$Department of Biostatistics, University of North Carolina at Chapel Hill, Chapel Hill, North Carolina, USA\\%
\today
}
\title{{Supporting Information for 'Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions' by Pedro L. Baldoni$^*$, Naim U. Rashid$^*$, and Joseph G. Ibrahim$^*$}}
} \fi

\if0\blind{
\date{%
\today
}
\title{{Supporting Information for 'Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions'}}
} \fi

@@

>>>>

\if1\blind{
\date{%
$^*$Department of Biostatistics, University of North Carolina at Chapel Hill, Chapel Hill, North Carolina, USA\\%
%\today
}
\title{{Supporting Information for 'Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions' by Pedro L. Baldoni$^*$, Naim U. Rashid$^*$, and Joseph G. Ibrahim$^*$}}
} \fi

\if0\blind{
\date{%
%\today
}
\title{{Supporting Information for 'Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions'}}
} \fi

@@

In the chromatin segmentation analysis, ChromHMM was run with the commands:
\begin{itemize}
\item ChromHMM-BinarizeBed: \textit{java -jar $\sim$/ChromHMM/ChromHMM.jar BinarizeBam -b 500 -c 'controldir' ~/ChromHMM/CHROMSIZES/hg19.txt 'inputbamdir' 'input.txt' 'outputbinarydir'} and
\item ChromHMM-LearnModel: \textit{unset DISPLAY \&\& java -jar $\sim$/ChromHMM/ChromHMM.jar LearnModel -p 0 -b 500 'outputbinarydir' 'outputdir' 'numstates' hg19},
\end{itemize}

such that 'controldir' is the directory with control experiments, 'inputbamdir' is the directory with input bam files, 'input.txt' is an experimental table, 'outputbinarydir' is the output directory with binarized files, 'outputdir' is the model output, and 'numstates' is the number of states in the model.

### 2021-04-02 Latex/letter/210305_letter/letter.Rmd
Update text

We thank you for the constructive comments regarding our paper "Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions".

Our revised paper has 20 pages in total, as requested, including appendices and references, and excluding the title page and abstract, figures and tables. We emphasize that the our presented method, previously called mixNBHMM, is now called epigraHMM. We refer to our method as epigraHMM throughout this revised version of our paper.

>>>>

We appreciate the constructive comments regarding our paper "Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions".

Our revised paper has 20 pages in total, as requested, including appendices and references, and excluding the title page and abstract, figures and tables. We emphasize that our presented method, previously called mixNBHMM, is now called epigraHMM. We refer to our method as epigraHMM throughout this revised version of our paper.

### 2021-04-02 Latex/paper/paper.tex
Update text

\begin{figure}
\begin{center}
\includegraphics[scale = 0.75]{../../Main/Figure_epigraHMM/Figure_epigraHMM.pdf}
\end{center}
\caption{At the top, differential peak calls from current methods (under false discovery rate control of 0.05) for the histone modification H3K27me3 among cell lines Helas3, Hepg2, and Huvec. Shaded regions indicate  observed differential enrichment, and each vertical line type bordering each region represents a different combinatorial pattern of enrichment across cell lines. Optimal methods would call broad peaks inside shaded regions and no peaks outside them. At the bottom, general overview of the presented method for simultaneous detection and classification of broad and short differential regions of enrichment. This figure appears in color in the electronic version of this article.
\label{fig:fig_motivation}}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[scale = 0.65]{../../Main/Figure_Simulation_SequencingReads/Figure_Simulation_SequencingReads.pdf}
\end{center}
\caption{Sequencing read-based simulation from the csaw pipeline. (A): average observed sensitivity and FDR for various methods. (B): scatter plot of average ratio of called and simulated peaks (y-axis) and number of called peaks intersecting true differential regions (x-axis). (C): box plot of computing time (in minutes) for various algorithms. (D): an example of differential peak calls under a nominal FDR control of 0.05. Shaded areas indicate true differential peaks. This figure appears in color in the electronic version of this article.
\label{fig:fig_simulation_csaw}}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[scale = 0.65]{../../Main/Figure_Broad/Figure_Broad.pdf}
\end{center}
\caption{Analysis of broad ENCODE data. (A): ROC curves of DPCs covering differentially enriched genes ($|\text{LFC}|>2$ of ChIP-seq counts) for H3K36me3. (C): sensitivity (y-axis) and average H3K36me3 differential peak size (kbp; x-axis) of various methods under different nominal FDR thresholds (observed FDR annotated next to data points). (B), (D), and (F): example of peak calls from H3K36me3, H3K27me3, and EZH2, respectively, under a nominal FDR control of 0.05. Posterior probabilities of the HMM differential state are shown at the bottom of each panel. (E): computing time and peak memory usage of genome-wide analysis from various methods. This figure appears in color in the electronic version of this article.
\label{fig:fig_broad}}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[scale = 0.65]{../../Main/Figure_Narrow/Figure_Narrow.pdf}
\end{center}
\caption{Analysis of short ENCODE data. (A) and (C): median LFC and correlation between cell lines of ChIP-seq counts from differential peaks for CTCF and H3K4me3, respectively. (B), (D), and (F): example of peak calls from CTCF, H3K4me3, and H3K27ac, respectively. Posterior probabilities of the HMM differential state are shown at the bottom of each panel. (E): computing time and peak memory usage of genome-wide analysis from various methods. Results are shown under a nominal FDR control of 0.05. This figure appears in color in the electronic version of this article.
\label{fig:fig_sharp}}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[scale = 0.7]{../../Main/Figure_MultiMarks/Figure_MultiMarks.pdf}
\end{center}
\caption{Genomic chromatin state segmentation and classification. (A): distribution of base pairs (y-axis) and the Viterbi sequence of states (x-axis). (B): estimated mixture probabilities and the associated differential combinatorial patterns. (C): choice of best model with two mixture components via BIC. (D): density estimate from expression of genes intersecting differential peaks associated with the enrichment of H3K36me3 alone or the enrichment of H3K27me3 and EZH2 in consensus. (E): cycle-regulated genes and their associated average mixture posterior probabilities associated with differential H3K36me3 enrichment pattern of overlapping differential peaks. (F): example of a genomic region with differential peaks and genes, colored according to their classification and expression levels, respectively. This figure appears in color in the electronic version of this article.
\label{fig:fig_multimarks}}
\end{figure}

To assess the biological significance of our results, we associated the distribution of transcripts per million (TPM, from matching RNA-seq experiments) of protein-coding genes with combinatorial patterns of overlapping differential peaks. To assign combinatorial patterns to differential peaks, we chose the combination pertaining to the most prevalent mixture component across windows by using the maximum estimated mixture model posterior probability, $Pr(W_{jl}=1|Z_{j}=2,\mb{y}_{..j},\mb{x};\bs{\Psi}^{(t)})$, $j=1,\ldots,M$. Genes associated with transcribed chromatin states had a significantly higher distribution of TPM counts than genes associated with repressed chromatin states (Figure~\ref{fig:fig_multimarks}D). In fact, nearly all known Helas3 cycle-regulated genes (therefore genes expressed at some stage of the cell life cycle; \cite{whitfield2002identification}) were associated with high average mixture posterior probabilities of differential enrichment for H3K36me3 alone (Figure~\ref{fig:fig_multimarks}E).

The expression of such genes are known to correlate with proliferative states of tumors and their study may help characterize their role in cancer \citep{whitfield2002identification}. Differential peaks called by epigraHMM, and their assigned differential combinatorial pattern, often agreed with the biological roles as well as the expression levels of associated genes (Figure~\ref{fig:fig_multimarks}F; results from ChromHMM with 3 states for comparison). epigraHMM offers the benefit of simultaneously detecting differential peaks and classifying differential combinatorial patterns of enrichment even in the context of genomic segmentation of highly diverse epigenomic marks (Web Appendix D4). By using the BIC for model selection, one can choose the number of biologically relevant mixture components to be included in the model, a task that may not be as straightforward in methods such as ChromHMM (Web Appendix B2).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\backmatter
% \section*{Acknowledgements}
% The authors thank Professor A. Sen for some helpful suggestions,
% Dr C. R. Rangarajan for a critical reading of the original version of the
% paper, and an anonymous referee for very useful comments that improved
% the presentation of the paper.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Supporting Information}
Web Appendices A-D, referenced in Sections 1-6, and a link to the implemented software are available with this paper at the Biometrics website on Wiley Online Library. MACS2 peak caller is available at https://github.com/macs3-project/MACS. epigraHMM has been implemented as an R package and it is available for download at https://github.com/plbaldoni/epigraHMM.

>>>>

\begin{figure}
\begin{center}
\includegraphics[scale = 0.75]{../../Main/Figure_epigraHMM/Figure_epigraHMM.pdf}
\end{center}
\caption{At the top, differential peak calls from current methods (under false discovery rate control of 0.05) for the histone modification H3K27me3 among cell lines Helas3, Hepg2, and Huvec. Shaded regions indicate  observed differential enrichment, and each vertical line type bordering each region represents a different combinatorial pattern of enrichment across cell lines. Optimal methods would call broad peaks inside shaded regions and no peaks outside them. At the bottom, general overview of the presented method for simultaneous detection and classification of broad and short differential regions of enrichment. This figure appears in color in the electronic version of this article, and any mention of color refers to that version.
\label{fig:fig_motivation}}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[scale = 0.65]{../../Main/Figure_Simulation_SequencingReads/Figure_Simulation_SequencingReads.pdf}
\end{center}
\caption{Sequencing read-based simulation from the csaw pipeline. (A): average observed sensitivity and FDR for various methods. (B): scatter plot of average ratio of called and simulated peaks (y-axis) and number of called peaks intersecting true differential regions (x-axis). (C): box plot of computing time (in minutes) for various algorithms. (D): an example of differential peak calls under a nominal FDR control of 0.05. Shaded areas indicate true differential peaks. This figure appears in color in the electronic version of this article, and any mention of color refers to that version.
\label{fig:fig_simulation_csaw}}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[scale = 0.65]{../../Main/Figure_Broad/Figure_Broad.pdf}
\end{center}
\caption{Analysis of broad ENCODE data. (A): ROC curves of DPCs covering differentially enriched genes ($|\text{LFC}|>2$ of ChIP-seq counts) for H3K36me3. (C): sensitivity (y-axis) and average H3K36me3 differential peak size (kbp; x-axis) of various methods under different nominal FDR thresholds (observed FDR annotated next to data points). (B), (D), and (F): example of peak calls from H3K36me3, H3K27me3, and EZH2, respectively, under a nominal FDR control of 0.05. Posterior probabilities of the HMM differential state are shown at the bottom of each panel. Shaded areas highlight differential peak regions. (E): computing time and peak memory usage of genome-wide analysis from various methods. This figure appears in color in the electronic version of this article, and any mention of color refers to that version.
\label{fig:fig_broad}}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[scale = 0.65]{../../Main/Figure_Narrow/Figure_Narrow.pdf}
\end{center}
\caption{Analysis of short ENCODE data. (A) and (C): median LFC and correlation between cell lines of ChIP-seq counts from differential peaks for CTCF and H3K4me3, respectively. (B), (D), and (F): example of peak calls from CTCF, H3K4me3, and H3K27ac, respectively. Posterior probabilities of the HMM differential state are shown at the bottom of each panel. Shaded areas highlight differential peak regions. (E): computing time and peak memory usage of genome-wide analysis from various methods. Results are shown under a nominal FDR control of 0.05. This figure appears in color in the electronic version of this article, and any mention of color refers to that version.
\label{fig:fig_sharp}}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[scale = 0.7]{../../Main/Figure_MultiMarks/Figure_MultiMarks.pdf}
\end{center}
\caption{Genomic chromatin state segmentation and classification. (A): distribution of base pairs (y-axis) and the Viterbi sequence of states (x-axis). (B): estimated mixture probabilities and the associated differential combinatorial patterns. (C): choice of best model with two mixture components via BIC. (D): density estimate from expression of genes intersecting differential peaks associated with the enrichment of H3K36me3 alone or the enrichment of H3K27me3 and EZH2 in consensus. (E): cell cycle-regulated genes and their associated average mixture posterior probabilities associated with differential H3K36me3 enrichment pattern of overlapping differential peaks. (F): example of a genomic region with differential peaks and genes, colored according to their classification and expression levels, respectively. This figure appears in color in the electronic version of this article, and any mention of color refers to that version.
\label{fig:fig_multimarks}}
\end{figure}

To assess the biological significance of our results, we associated the distribution of transcripts per million (TPM, from matching RNA-seq experiments) of protein-coding genes with combinatorial patterns of overlapping differential peaks. To assign combinatorial patterns to differential peaks, we chose the combination pertaining to the most prevalent mixture component across windows by using the maximum estimated mixture model posterior probability, $Pr(W_{jl}=1|Z_{j}=2,\mb{y}_{..j},\mb{x};\bs{\Psi}^{(t)})$, $j=1,\ldots,M$. Genes associated with transcribed chromatin states had a significantly higher distribution of TPM counts than genes associated with repressed chromatin states (Figure~\ref{fig:fig_multimarks}D). In fact, nearly all known Helas3 cell cycle-regulated genes (therefore genes expressed at some stage of the cell cycle; \cite{whitfield2002identification}) were associated with high average mixture posterior probabilities of differential enrichment for H3K36me3 alone (Figure~\ref{fig:fig_multimarks}E).

The expression of such genes are known to correlate with proliferative states of tumors and their study may help characterize their role in cancer \citep{whitfield2002identification}. Differential peaks called by epigraHMM, and their assigned differential combinatorial pattern, often agreed with the biological roles as well as the expression levels of associated genes (Figure~\ref{fig:fig_multimarks}F; results from ChromHMM with 4 states for comparison). epigraHMM offers the benefit of simultaneously detecting differential peaks and classifying differential combinatorial patterns of enrichment even in the context of genomic segmentation of highly diverse epigenomic marks (Web Appendix D4). By using the BIC for model selection, one can choose the number of biologically relevant mixture components to be included in the model, a task that may not be as straightforward in methods such as ChromHMM (Web Appendix B2).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\backmatter

\section*{Acknowledgements}
The authors wish to thank the editor, associate editor and three referees for helpful comments and suggestions, which have led to an improvement of this article. This research was partially supported by NIH grants GM70335, P01CA142538, P30CA016086, and P50CA058223 and by the Brazilian funding agency CAPES (13195/13‐1).

\section*{Data Avalability Statement}

The data that support the findings of this paper are openly available in the ENCODE Project portal \citep{sloan2016encode} at https://www.encodeproject.org.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Supporting Information}
Web Appendices A-D, referenced in Sections 1-6, and a link to the implemented software are available with this paper at the Biometrics website on Wiley Online Library. MACS2 peak caller is available at https://github.com/macs3-project/MACS. epigraHMM has been implemented as an R package and it is available for download at https://github.com/plbaldoni/epigraHMM. All of the scripts used in the analyses presented in
this paper can be downloaded at https://github.com/plbaldoni/epigraHMMPaper.
